Verification of Stencil Computations with Spatial Specifications

### 2016-11-02 paper.tex
POPL version
starting on PLDI version

\title{Verification of Stencil Computations with Spatial
  Specifications}
\titlebanner{DRAFT -- Do not distribute}
%\authorinfo{}{}{}
\authorinfo{Dominic Orchard \and Mistral Contrastin
\and Matthew Danish \and Andrew Rice}{University of Cambridge}{}

@@

\begin{abstract}
  Verifying the correctness of numerical computations relies on first
  giving a specification of their behaviour. In many cases, a
  numerical computation is derived from an underlying mathematical
  model, but this is often infeasible as a specification since it
  is highly decoupled from the implementation. In this paper, we consider a 
  lightweight specification technique between low-level and
  high-level specifications for numerical code. We
  target \emph{stencil computations}---a common idiom in numerical
  code, but one which is error-prone due to fine-grained indexing
  errors. We show that in practice, stencil computations tend
  to have a regular shape which is amenable to static analysis and
  simple specification. We describe an abstract spatial specification
  language for stencil computations that can be embedded as
  annotations in source code, along with inference, checking, and
  specification synthesis. We evaluate our
  language against a corpus of numerical Fortran code, and show that
  the vast majority of stencil computations indeed have a simple,
  regular, static shape. Around 10\% of the stencils found evoke a
  programming pattern for which indexing errors are plausible. 
  We simulate relevant programming errors in two stencil examples, 
  and show that we are able to detect the vast majority of potential
  errors. This paper details our stencil specification language, a
  denotational model, inference,
  checking, and synthesis procedures, our implementation, and
  our evaluation studies.
\end{abstract}

\keywords{program verification, specification, stencils}

\bibliographystyle{abbrvnat}

\section{Introduction}

\emph{Stencils} are a ubiquitous programming pattern, common in
scientific and numerical computing applications. Informally, a stencil
computation outputs an array, where the value at each index $i$ of
this array is calculated from a \emph{neighbourhood} of values around $i$ in
some input array(s), \eg{}, the Game of Life, convolutions in image
processing, approximations to differential equations. For example, the
following computes the one-dimensional discrete Laplace
transform (an approximation to a derivative) in Fortran:
%
\begin{minted}{fortran}
do iter = 0, itermax
   do i = 1, (n-1)
      b(i) = a(i-1) - 2*a(i) + a(i+1)
   end do
   a = b
end do
\end{minted}
%
Line 3 is the core of the stencil computation, calculating
values at \texttt{b(i)} from a neighbourhood of elements about
\texttt{a(i)}. Line 5 swaps
\texttt{a} and \texttt{b} between iterations, where \texttt{b} becomes the
input for the next iteration. %In Fortran, parentheses \texttt{( )} are used
%for array subscripts rather than the more familiar bracket syntax \texttt{[ ]}.
%
%This stencil computation exhibits statically decideable
%spatial and temporal relationships between $\texttt{a}$ and
%$\texttt{b}$.
The shape of the data access pattern on line 3 determines other aspects of the
program and its efficient implementation: how much ``boundary'' is
needed for the array, the most cache-efficient layout in memory,
and the partitioning shape for parallel implementations.

In this example, the access pattern is simple, static, and easily
understood. More complex stencil computations
are much more prone to error from simple lexical mistakes. For example,
Figure~\ref{ref:navier-stokes-fragment} shows three lines from a
Navier-Stokes fluid simulator in which two arrays are read with
different data access patterns, across two dimensions. The interaction
is much harder to understand, with the potential for the developer to
introduce an error via simple textual mistakes, for
example writing $\texttt{(i-1,j)}$ instead of $\texttt{(i+1,j)}$.

In practice, most stencil computations have a regular shape that can
be described simply and abstractly with a small set of coarse-grained
spatial descriptions. We introduce a simple specification language of
such descriptions, abstracting over the fine-grained detail of stencil
access patterns. For the Laplace example, our
inference procedure provides the following specification to line 3:
%
\begin{minted}[linenos=false]{fortran}
!=  stencil centered(depth=1, dim=1) :: a
\end{minted}
%
This describes that \texttt{a} is accessed with a
symmetrical pattern (``centered'') to a depth of one in each
direction in its first (and only) dimension.
%The second line explains
%the temporal relationship between \texttt{b} and \texttt{a}: that
%previous time step for \texttt{b} is actually provided by \texttt{a},
%and vice versa. This is explained as a mutual dependence between
%\texttt{a} and \texttt{b}.
The inferred specification for the Navier-Stokes example is shown 
in Fig.~\ref{ref:navier-stokes-fragment}(b). The
specification requires that, over the whole fragment, \texttt{u} is
accessed with a centered pattern to depth of 1 in both dimensions
(this is known as the \emph{five-point stencil}) and \texttt{v}
is accessed in a neighbourhood bounded by forward to depth of $1$ in
the first dimension and backward to a depth of $1$ in the second
dimension. Fig.~\ref{ref:navier-stokes-fragment}(c) illustrates
the specification pictorially.

In this paper, we make the following contributions:
%
\begin{itemize}[leftmargin=0.8em]
\item we introduce a specification language for stencils 
  that captures many common forms of data access pattern
  (Section~\ref{sec:lang});

\item we detail inference, checking, and specification synthesis 
algorithms for our language (Section~\ref{sec:analysis}), derived from
a denotational model for the language (Section~\ref{sec:semantics});

\item we report on a quantitative study of
  stencil computations on a corpus of numerical Fortran programs
  (ranging from small to large),  totalling one million lines of code. Our tool identifies and
  synthesises specifications for 35,000 stencil computations in the
  corpus. Approximately 10\% of the stencils we found are non-trivial,
  corresponding to code which is a possible source of errors;

\item we give a detailed verification case study of two
  particular stencil computation, simulating programming errors.
  Our approach is able to detect the vast majority of possible
  indexing errors.
\end{itemize}
%

\begin{figure}[t]
\begin{minted}[firstnumber=20,xleftmargin=2em]{fortran}
du2dx = ((u(i,j)+u(i+1,j))*(u(i,j)+u(i+1,j))+   &
  gamma*abs(u(i,j)+u(i+1,j))*(u(i,j)-u(i+1,j))- &
  (u(i-1,j)+u(i,j))*(u(i-1,j)+u(i,j))-          &
  gamma*abs(u(i-1,j)+u(i,j))*(u(i-1,j)-u(i,j))) &
  /(4.0*delx)

@@

f(i,j) = u(i,j)+del_t*(laplu/Re-du2dx-duvdy)
\end{minted}
(a) Excerpt of a Navier-Stokes fluid simulator
(based on~\citet{griebel1997numerical}) showing highly-detailed stencil computation.
\vspace{0.5em}
\hrule
%
\begin{minted}[linenos=false,xleftmargin=1em]{fortran}
!= stencil centered(depth=1,dim=1)*reflexive(dim=2) + centered(depth=1,dim=2)*reflexive(dim=1) :: u

!= stencil forward(depth=1,dim=1) * backward(depth=1,dim=2) :: v
\end{minted}
(b) Inferred and synthesised specification for (a) by CamFort.
\vspace{0.5em}
\hrule
%
\begin{center}
\begin{tikzpicture}[scale=0.85]
\node at (1.25,1.22) {i, j};
\draw (1,1) rectangle (1.5,1.5);
\draw (1.5,1) rectangle (2,1.5);
\draw (1,0.5) rectangle (1.5,1);
\draw (0.5,1) rectangle (1,1.5);
\draw (1,1.5) rectangle (1.5,2);
%
\node at (4.25,1.22) {i, j};
\draw (4,1) rectangle (4.5,1.5);
\draw (4.5,1) rectangle (5,1.5);
\draw (4.5,1.5) rectangle (5,2);
\draw (4,1.5) rectangle (4.5,2);
\end{tikzpicture}
\end{center}
\vspace{-0.5em}
(c) Pictorial representation of the two stencil specifications.
The horizontal dimension is \texttt{dim=1} and the vertical is \texttt{dim=2}.
\caption{Fragment of a Navier-Stokes fluid simulator and its
  specification in our language.}
\label{ref:navier-stokes-fragment}
\vspace{-0.8em}
\end{figure}

%This paper considers the technical aspects of the language,
%its semantics, inference, checking, and synthesis procedures.
%We also evaluate our implementation on real scientific code 
%used actively in research, ranging from small ($\sim$700loc) to large
%($\sim$600kloc). 
 %Our case studies include the Unified Weather 
%model used by the UK Met Office~\cite{um-website} ($\approx{}600\textit{kloc}$) and other
%research-based numerical models, ranging from small to large.

\section{Stencil specification language}
\label{sec:lang}

Our specification language is based on the hypothesis
that most forms of array access in numerical code have
a fixed, statically-determined access pattern. For example, the
\emph{five-point stencil} on a two-dimensional array reads from array
indices $(i, j)$, $(i-1, j)$, $(i+1, j)$, $(i, j-1)$, and $(i, j+1)$
for all $i, j$ within the inner boundary of the array (to avoid
out-of-bounds access at the edges). We revisit this hypothesis
in Section~\ref{sec:evaluation} with the inference of
such regular stencil patterns on our corpus of numerical programs.

Section~\ref{sec:syntax} outlines the syntax.
 Section~\ref{sec:eqs} provides an equational
theory for specifications and a notion of
sub-specification.  Section~\ref{sec:semantics}
gives a semantic model via a set-based interpretation, for which the
equality and sub-specification relation are proven sound.

\paragraph{Notation and convention}
%\label{sec:notation}

\renewcommand*{\arraystretch}{0.8}
For the target language, $e$ ranges over expressions (may be impure) and $v$
over its (imperative) variables. %By
%``variable'', we mean in the imperative sense (named binders to memory cells).

\begin{definition}[Base induction variable]
  A integer variable \textit{v} is a \emph{base induction
    variable} if it is the control variable of a ``for''
  loop,\footnote{Or equivalent, \eg{}, \texttt{do} in Fortran, for our implementation.} incremented by $1$ per
iteration. The variable is marked as an induction variable
only within the body of the loop. Since we ignore non-base
(derived) induction variables, we simply say \emph{induction
  variable}. Induction variables are ranged over by $i$, $j$, $k$ throughout. 
\end{definition}

\begin{definition}[Array subscript and index]
  An \emph{array subscript} is an expression that indexes an array, which we denote
  by $v(\bar{e})$ for the source language where $\bar{e}$ is shorthand
  for a syntactic list of indexing expressions. A \emph{relative
    index} is a list $\bar{e}$ where each $e \in \bar{e}$ is defined
  in terms of base induction variables.
\end{definition}

\begin{definition}[Neighbourhood index]
  For an array subscript $v(\bar{e})$, we say that $e \in \bar{e}$
  is a \emph{neighbourhood index} if it is of the form
  $e \equiv i$, $e \equiv i + a$ or $e \equiv i - a$ where $a$ is an
  integer constant. That is, a neighbourhood index is a
  relative index defined as constant translations of base induction
  variables. (The relation $\equiv$ here identifies terms up-to commutativity
  of $+$ and the inverse relation of $+$ and $-$ \eg{},
  $(-b) + i \equiv i - b$).  We classify neighbourhood indices 
  by the predicate $\neigh_I$ where $I$ is a set of induction
  variables. We omit $I$ when clear from the context.
\label{def:neighbour}
\end{definition}

\subsection{Specification syntax}
\label{sec:syntax}

Figure~\ref{fig:syntax} gives the syntax of stencil specifications,
which we introduce in stages below.  The top-level is given by the
\nonterm{specification} production which splits into either a
\nonterm{regionDec} (region declaration) or a \nonterm{specDec}
(specification declaration). Specification declarations associate
(via the operator $\texttt{::}$) a specification to one or more program variables
$\bar{v}$, describing how the array variables $\bar{v}$ are accessed.
%Specification declarations are either \nonterm{spatial} or
%\nonterm{temporal} specifications and describe, for spatial
%specifications,
%Specifications declarations describe
%how an array variable $v$ is read
%, and for temporal
%specifications what variables are used to define the array variable $v$.
%%

%\subsubsection{Spatial specification syntax}

\emph{Regions} are the central building blocks of spatial
specifications. Regions can either be declared alone with a
\nonterm{regionDec}, assigning a region specification \nonterm{region} to
a region variable \nonterm{rvar} (via $\texttt{::}$) or given directly within a
\nonterm{spatial} specification.

\paragraph{Region constants}

Regions have as terminals the \emph{region constants}, denoted by
\term{reflexive}, \term{forward}, \term{backward}, or \term{centered}. Each
region constant except \term{reflexive} is given a depth parameter $n$ (natural
number greater than 0). Every region constant receives a dimension identifier $d$ (also a
natural number greater than 0). Region constants specify that an array is read
from by a collection of indices which in the $d^{th}$ dimension are
neighbourhood indices ranging from $i + 0$ up to $i \pm n$ inclusively,
\eg{}, the following is a valid stencil, reading from \term{a} and writing to
\term{b}:
%%
\begin{minted}{fortran}
!= stencil forward(depth=2, dim=1) :: a
b(i, 0) = a(i, 0) + a(i+1, 0) + a(i+2, 0)
\end{minted}
%%
where \texttt{i} is an induction variable.  The specification
is associated to the array variable \texttt{a}. Note that the second
dimension is constant.

The \term{forward},
\term{backward}, and \term{centered} regions can all receive an
additional optional attribute \term{irreflexive} which marks
that the region does not include the origin (offset 0).
For example, the following is a \term{backward} stencil, which is similar
to \term{forward} but has negative neighbourhood indices, and which
is irreflexive:
%
\begin{minted}{fortran}
!= stencil backward(depth=2, dim=1, irreflexive) :: a
b(i) = a(i-1) + a(i-2)
\end{minted}
%
The \term{irreflexive} attribute cannot be used on \term{reflexive}
regions.

A \texttt{centered} stencil is a combination of \texttt{forward}
and \texttt{backward} stencils to the same depth, \eg{},
%%
\begin{minted}{fortran}
!= stencil centered(depth=1, dim=1) :: a
b(i) = ( a(i) + a(i-1) + a(i+1) ) / 3.0
\end{minted}

\begin{figure}[t]
\vspace{-0.5em}
\begin{align*}
\def\arraystretch{1.2}
\setlength{\arraycolsep}{0.2em}
\newcommand{\dimTy}{\mathbb{N}_{>0}}
\begin{array}{rl}
\nonterm{specification} ::= & \nonterm{regionDec} \mid \nonterm{specDec} \\
\nonterm{specDec} ::= & \term{stencil} \; \nonterm{spec} \;
                        \texttt{::} \; v \\
\nonterm{regionDec} ::= &  \texttt{region} \; \texttt{::} \; \nonterm{rvar} \; \texttt{=} \;
                         \nonterm{region}\\[0.4em]
%\nonterm{spec} ::= & \nonterm{spatial} \mid \nonterm{temporal}
%\\[1em]
\nonterm{spec} ::= & [\nonterm{approx},] \; [\nonterm{mult},] \; \nonterm{region} \\
\nonterm{mult} ::= & \term{readOnce} \\
\nonterm{approx} ::= & \term{atMost} \; \mid \; \term{atLeast} \\[0.1em]
\nonterm{region} ::= & \nonterm{rvar} \; \\
\multicolumn{2}{l}{\qquad\qquad \mid \; \stenRefl{\dimTy}} \\
\multicolumn{2}{l}{\qquad\qquad \mid \;\stenFwd{\mathbb{N}_{>0}}{\dimTy}{\;[, \texttt{irreflexive}]}} \\
\multicolumn{2}{l}{\qquad\qquad \mid \; \stenBwd{\mathbb{N}_{>0}}{\dimTy}{\;[, \texttt{irreflexive}]}} \\
\multicolumn{2}{l}{\qquad\qquad \mid \; \stenCen{\mathbb{N}_{>0}}{\dimTy}{\;[, \texttt{irreflexive}]}} \\
\multicolumn{2}{l}{\qquad\qquad \mid \; \nonterm{region} \; \term{+}
  \; \nonterm{region} \; \mid \; \nonterm{region} \; \term{*} \; \nonterm{region}} \\[0.5em]
%\multicolumn{2}{l}{\qquad\qquad \mid \; \nonterm{rvar}}  
%\\[0.5em]
%\nonterm{temporal} ::= \; & \term{dependency} \; (v \; \{ , v \}) [, \texttt{mutual}]
%  \\[0.5em]
\nonterm{rvar} ::= \; & [\text{\term{a}-\term{z}$\,$\term{A}-\term{Z}$\,$\term{0}-\term{9}}]+\\[-0.5em]
\end{array}
\end{align*}
\caption{Specification syntax (EBNF grammar)}
\label{fig:syntax}
\vspace{-0.8em}
\end{figure}

\paragraph{Sum and product of regions}
%%
Region terms can be combined using the sum operator
\term{+} or the product operator \term{*}.

The product of two regions $r \texttt{*} r'$ specifies that an array
is read with neighbourhood indices drawn from the bounding box created
by the regions two regions $r$ and $r'$. For example, the following
code (defining a \emph{nine-point stencil}) has a specification given by the
product of two \texttt{centered} regions in each dimension:
%%
\begin{minted}[breakindent=2.9em]{fortran}
x = a(i, j)   + a(i-1, j)   + a(i+1, j)
y = a(i, j-1) + a(i-1, j-1) + a(i+1, j-1)
z = a(i, j+1) + a(i-1, j+1) + a(i+1, j+1)
!= stencil centered(depth=1, dim=1) * centered(depth=1, dim=2) :: a
b(i, j) = (x + y + z) / 9.0
\end{minted}
%%
This pattern is common in image convolution applicaitons (for example
for edge detection). The specification ranges over the
values that flow to the array subscript on the left-hand side,
and so ranges over the intermediate assignments to \term{x},
\term{y}, and \term{z}.

The region constructor \texttt{*} can also be interpreted as a
kind of conjunction on specifications since they constrain
indices to be within both regions simultaneously.

The sum of two regions $r \texttt{+} r'$ specifies that an array is
read using the neighbourhood indices described by either $r$ or
$r'$; it acts as a disjunction on specifications. For
example, the following gives the specification of a five-point
stencil which is the sum of two compound \texttt{reflexive} and
\texttt{centered} regions in each dimension:
%%
\begin{minted}{fortran}
!= stencil centered(depth=1, dim=1)*reflexive(dim=2) + centered(depth=1, dim=2)*reflexive(dim=1) :: a
b(i,j) = -4*a(i,j) + a(i-1,j) + a(i+1,j) &
                   + a(i,j-1) + a(i,j+1)
\end{minted}
%%
Here the left-hand side of \texttt{+} says that when the second dimension
(induction variable $j$) is fixed at the origin, the first dimension
(induction variable $i$) accesses the immediate vicinity of the origin
(to depth of one). The right hand side of \texttt{+} is similar but the dimensions are reversed.
This reflects the symmetry under rotation of the five-point stencil.

As another example, the following
implements and specifies the \emph{Roberts cross}
edge-detection convolution~\cite{davis1975survey}:
%%
\begin{minted}[breakindent=5.5em]{fortran}
do j=0, jmax-1
  do i=0, imax-1
    x = a(i,j)-a(i+1,j+1); y = a(i+1,j)-a(i,j+1)
    != stencil forward(depth=1,dim=1) * forward(depth=1,dim=2) :: a
    b(i,j) = sqrt(x*x+y*y)
  end do
end do
\end{minted}

\paragraph{Region declarations and variables}

Region specifications can be assigned to region variables via
a region declaration, which can be used later to form a spatial
specification. For example, the specification from the previous
 example can be restated as:
%%
\begin{minted}{fortran}
!= region :: r1 = forward(depth=1, dim=1)
!= region :: r2 = forward(depth=1, dim=2)
!= region :: robertsCross = r1*r2
!= stencil robertsCross :: a
\end{minted}
This is especially useful for common stencil patterns, such as Roberts cross,
as the region can be defined once and reused.
%%
\paragraph{Modifiers}
%%
Region specifications can be modified 
by \emph{approximation} and \emph{multiplicity} information 
(in the \textit{spec} rule of Fig.~\ref{fig:syntax}).

The \texttt{readOnce} modifier enforces that no index appears more
than once (that is, its multiplicity is one). For example, in all of
the previous examples the \texttt{readOnce} modifier could be added,
\eg{}
%
\begin{minted}{fortran}
!= stencil readOnce, backward(depth=2, dim=1) :: a
b(i+1) = a(i) + a(i-1) + a(i-2)
\end{minted}
%
This specification would be invalid if any of the
array subscripts were repeated. This modifier provides a way to
rule out any accidental repetition of array subscripts.
The notion is similar to that of \emph{linearity} in linear type
systems~\cite{wadler1990linear}, where a value must be used
exactly once. We opt for the more informative and easily understood name
\texttt{readOnce}. This modifier is optional, so it need not
be present even if the stencil is linear.

%At the region level, \texttt{irreflexive} modifier qualifies base
%regions of \texttt{forward}, \texttt{backward}, and \texttt{centered}.
%It indicates that in the dimension of the region there is no neighbour index
%with $0$ offset, \eg{},
%
%\begin{minted}{fortran}
%!= stencil centered(depth=1, dim=1, irreflexive) :: a
%b(i) = a(i+1) + a(i-1)
%\end{minted}
%
%or with just a \texttt{forward} region an example would be:
%%
%\begin{minted}{fortran}
%!= stencil forward(depth=1, dim=1, irreflexive) :: a
%b(i) = a(i+1)
%\end{minted}

%\paragraph{Under and over-approximation}

In some cases, it is useful to give a lower and/or upper bound for a
stencil. This can be done using either the \term{atMost} or
\term{atLeast} modifiers. This is particularly useful in situations
where there is a non-contiguous stencil pattern, which cannot be expressed
precisely in our specification syntax. For example:
%%
\begin{minted}{fortran}
!= stencil atLeast, reflexive(dim=1)       :: a
!= stencil atMost, forward(depth=2, dim=1) :: a
b(i) = a(i) + a(i+2)
\end{minted}
%%
%\subsubsection{Temporal specifications}

\paragraph{Note on the design}

The names ``forward'', ``backward'' and ``centered''
are inspired by standard terminology in numerical analysis
for the shape of discretisation schemes. For example,
the standard \emph{explicit method} for approximating
PDEs is known as the \emph{Forward Time, Centered Space} (FTCS)
scheme~\cite{dawson1991finite}. For
the one-dimensional heat equation, an FTCS discretisation
provides code with the following stencil~\cite{recktenwald2004finite}:
%%
\begin{minted}{fortran}
do i=2, n-1
  u(i) = r*v(i-1) + r2*v(i) + r*v(i+1)
end do
\end{minted}
%%
where \texttt{r} and \texttt{r2} are constants.
A valid specification for this is:
%%
\begin{minted}{fortran}
!= stencil centered(depth=1, dim=1) :: v
\end{minted}
%%
Such a specification can be inferred, or can be inserted by a user
and checked against the code.

\subsection{Equational theory and sub-specifications}
\label{sec:eqs}

\begin{figure}
%\vspace{-0.2em}
\begin{align*}
\setlength{\arraycolsep}{0.05em}
\begin{array}{c}
%\framebox{$\textit{region} \equiv \textit{region}'$} \\[0.6em]
\begin{array}{rll}
(\textsc{F\, \texttt{+} \,F}) \;\; &
\stenFwdSR{n \, \textsf{max} \, m}{d}{r \vee s} \\
 \equiv \; & \stenFwdSR{n}{d}{r} \; \texttt{+} \; \stenFwdSR{m}{d}{s} \\[0.6em]
%
%
(\textsc{B\, \texttt{+} \,B}) \;\; &
\stenBwdSR{n \, \textsf{max} \, m}{d}{r \vee s} \\
 \equiv \; & \stenBwdSR{n}{d}{r} \; \texttt{+} \; \stenBwdSR{m}{d}{s} \\[0.6em]
%
%
(\textsc{C\, \texttt{+} \,C}) \;\; &
\stenCenSR{n \, \textsf{max} \, m}{d}{r \vee s} \\
\equiv \; & \stenCenSR{n}{d}{r} \; \texttt{+} \; \stenCenSR{m}{d}{s} \\[0.6em]
%
%
(\textsc{C\, \texttt{+} \,F}) \;\; & \stenCenSR{n}{d}{r \vee s} \\
\textit{$m \leq n$} \; \equiv \; & \stenCenSR{n}{d}{r} \; \texttt{+} \;
                      \stenFwdSR{m}{d}{s} \\[0.6em]
%
%
(\textsc{C\, \texttt{+} \,B}) \;\; &
\stenCenSR{n}{d}{r \vee s} \\
\textit{$m \leq n$} \; \equiv \; & \stenCenSR{n}{d}{r} \; \texttt{+} \;
                      \stenBwdSR{m}{d}{s} \\[0.6em]
%
%
(\textsc{B\, \texttt{+} \,F}) \;\; &
\stenCenSR{n}{d}{r \vee s} \\
\equiv \; & \stenFwdSR{n}{d}{r} \; \texttt{+} \; \stenBwdSR{n}{d}{s}
  \\[0.6em]
%
%
(\textsc{R\, \texttt{+} \,F}) \;\; &
\stenFwdS{n}{d} \\
\equiv \; & \stenReflSR{d} \; \texttt{+} \; \stenFwdSR{n}{d}{r} \\[0.6em]
%
%
(\textsc{R\, \texttt{+} \,B}) \;\; &
\stenBwdS{n}{d} \\
\equiv \; & \stenReflSR{d} \; \texttt{+} \; \stenBwdSR{n}{d}{r} \\[0.6em]
%
%
(\textsc{R\, \texttt{+} \,C}) \;\; &
\stenCenS{n}{d} \\
\equiv \; & \stenReflSR{d} \; \texttt{+} \; \stenCenSR{n}{d}{r} \\[0.6em]
%
%
(\textsc{irrefl}) \;\; & (R \, \term{*} \, S^\irreflS) \, \term{+} \,
                         (R^\irreflS \, \term{*} \, S) \;
\equiv \; R \, \term{*} \, S \\[0.4em]
%
%
(\texttt{+}\textsc{idem}) \;\; & S \; \texttt{+} \; S \; \equiv \; S \\[0.4em]
%
(\texttt{+}\textsc{comm}) \;\; & S \; \texttt{+} \; T \; \equiv \; T \;
                       \texttt{+} \; S \\[0.4em]
%
(\texttt{+}\textsc{assoc}) \;\; & R \; \texttt{+} \; (S \; \texttt{+} \; T) \; \equiv \; (R \;
                       \texttt{+} \; S) \; \texttt{+} \; T \\[0.4em]
(\texttt{*}\textsc{comm}) \;\; & S \; \texttt{*} \; T \; \equiv \; T \;
                       \texttt{*} \; S \\[0.4em]
%
(\texttt{*}\textsc{assoc}) \;\; & R \; \texttt{*} \; (S \; \texttt{*} \; T) \; \equiv \; (R \;
                       \texttt{*} \; S) \; \texttt{*} \; T \\[0.4em]
(\textsc{dist}) \;\; & R \; \texttt{*} \; (S \; \texttt{+} \; T) \; \equiv \; (R \;
                       \texttt{*} \; S) \; \texttt{+} \; (R
                       \; \texttt{*} \; T)
\end{array}
%
%\framebox{${\textit{specDec} :: \texttt{v}} \equiv
%{\textit{specDec'} :: \texttt{v}}$} \\[0.7em]
%\begin{array}{rl}
% MUTUAL
%(\textsc{mutual}) \; &
%\texttt{stencil} \; \texttt{dependency} (\texttt{v}), \texttt{mutual}
%  :: \texttt{w}
%\\
%\equiv \; & \texttt{stencil} \; \texttt{dependency} (\texttt{w}), \texttt{mutual} ::
%  \texttt{v} \\[0.5em]
%(\textsc{coalesce}) \; &
%\texttt{stencil} \; \texttt{dependency} (\bar{v})
%  :: \texttt{v} \\
%& ; \texttt{stencil} \; \texttt{dependency} (\bar{w})
%  :: \texttt{v}
%\\
%\equiv \; & \texttt{stencil} \; \texttt{dependency} (\bar{v}, \bar{w}) ::
%  \texttt{v} \\[0.5em]
%\end{array}
\end{array}
\end{align*}
\caption{Equations on specifications (\emph{regions})}
\vspace{-0.5em}
\label{fig:equations}
\end{figure}

Figure~\ref{fig:equations} lists the equational theory for
specifications given by the binary relation $\equiv$
on \emph{region} terms. The equations use abbreviations
\term{refl}, \term{fwd}, \term{bwd}, \term{cen}, and \irreflS{}
for \term{reflexive}, \term{forward}, \term{backward},
\term{centered}, and \irrefl{} respectively. Furthermore, the abbreviations include syntactic sugar
for quantifying over the absence or presence of the \irrefl{}
attribute on region constants as a boolean. For example,
\term{fwd} syntactic sugar is defined:
%%
\begin{align*}
\hspace{-0.25em}
& \stenFwdSR{n}{d}{r}
\; := \; \\[-0.4em]
& \qquad \begin{cases}
\stenFwd{n}{d}{, \irrefl} & \;\;\textit{if} \;\, \neg \, r \\[-0.1em]
\stenFwd{n}{d}{}  & \;\;\textit{if} \;\,  r
\end{cases} \\
\textit{or} \;\;
& \stenFwdS{n}{d} := \stenFwd{n}{d}{}
\end{align*}
%
and similarly for \term{bwd} and \term{cen}.

>>>>

\title{Verification of Stencil Computations with Spatial
  Specifications}
\titlebanner{DRAFT -- Do not distribute}
\authorinfo{}{}{}

@@

\emph{Stencils} are a ubiquitous programming pattern, common in
scientific and numerical computing applications. Informally, a stencil
computation computes an array whose value at each index $i$ is
calculated from a \emph{neighbourhood} of values around $i$ in some
input array(s), \eg{}, the Game of Life, convolutions in image
processing, approximations to differential equations. For example, the
following computes the one-dimensional discrete Laplace transform (an
approximation to a derivative) in Fortran:
%
\begin{minted}{fortran}
do i = 1, (n-1)
   b(i) = a(i-1) - 2*a(i) + a(i+1)
end do
\end{minted}
%
Values \texttt{b(i)} are calculated from a neighbourhood of elements
around \texttt{i} in the input array \texttt{a}.  In this example, the
access pattern is simple and easily understood. More complex
stencil computations are more prone to errors from simple lexical
mistakes. For example, the following is a three-line snippet from a
Navier-Stokes fluid simulator in which two arrays are read with
different data access patterns, across two dimensions.
%%
\begin{minted}[firstnumber=20,xleftmargin=4em]{fortran}
du2dx = ((u(i,j)+u(i+1,j))*(u(i,j)+u(i+1,j))+   &
  gamma*abs(u(i,j)+u(i+1,j))*(u(i,j)-u(i+1,j))- &
  (u(i-1,j)+u(i,j))*(u(i-1,j)+u(i,j))-          &
  gamma*abs(u(i-1,j)+u(i,j))*(u(i-1,j)-u(i,j))) &
  /(4.0*delx)

@@

f(i,j) = u(i,j)+del_t*(laplu/Re-du2dx-duvdy)
\end{minted}
%%
The access pattern is much harder to understand than the
one-dimensional Laplace.  The miasma of indexing expressions of the
form \texttt{var(i$\pm$a, j$\pm$b)} is not only hard to read, but is
prone to simple textual input mistakes, \eg{}, swapping \texttt{-} and
\texttt{+}, missing an indexing term, or transforming the wrong
variable \eg{} \texttt{(i+1,j)} instead of \texttt{(i,j+1)}.

In practice, we find that the usual development procedure for complex
stencil computations involves some ad hoc testing to ensure that no
transcription mistakes have been made. This may involve visual
inspection of the results, based on a rendered image. Typically such
testing is then discarded once the code is seen to be correct.

This is not the only information that is typically discarded. The
shape of the indexing pattern is often a result of choices made in the
numerical-analysis discretisation procedure used to convert a set of
continuous equations into a discrete approximation. Rarely are these
decisions captured in the source code, yet the shape of access is
usually uniform and is given a perspicuous and concise description in
numerical analysis literature \eg{}, ``\emph{centered in space, to a
  depth of 1}'' referring to indexing terms \texttt{a(i)},
\texttt{a(i-1)} and \texttt{a(i+1)}.

To support the development of correct stencil computations, we propose
a simple, abstract specification language for the data access pattern
of stencils. The shape of the stencil computation in the
Laplace example is specifed in our language as:
%
\begin{minted}[linenos=false]{fortran}
!=  stencil centered(depth=1, dim=1) :: a
\end{minted}
%
This describes that \texttt{a} is accessed with a symmetrical pattern
(``centered'') to a depth of one in each direction in its first
dimension.  The stencil computation shape in the Navier-Stokes example
is specified as:
%
\begin{minted}[linenos=false,xleftmargin=4em,breakindent=3em]{fortran}
!= stencil centered(depth=1,dim=1)*reflexive(dim=2) + centered(depth=1,dim=2)*reflexive(dim=1) :: u

!= stencil forward(depth=1,dim=1) * backward(depth=1,dim=2) :: v
\end{minted}
%
The specification requires that, over the whole fragment, \texttt{u}
is accessed with a centered pattern to depth of 1 in both dimensions
(this is known as the \emph{five-point stencil}) and \texttt{v} is
accessed in a neighbourhood bounded forwards to depth of $1$ in the
first dimension and backward to a depth of $1$ in the second
dimension.

We provide a verification tool for such specification which can check
the correctness of code against the specification, and can also infer
specifications, synthesising these as comments in the
source code. The main features and benefits of our approach are as follows:
%
\begin{itemize}
\item The main use case is that a programmers writes a specification
  first (perhaps informed by their numerical analysis), then writes
  the corresponding source code, and lastly has the tool check its
  conformance to the specifications. This specify-and-check approach
  removes the usually ad hoc testing effort, and provides the
  opportunity for easy re-verification in the event of any refactoring
  of software maintenance activities.
%
\item The specification provides a concise description which can be
  preserved in the source code to provide documentation to future
  developers, contrasting with ad hoc tests which are usually
  discarded or specialised to the development environment. Separate from the
  verification procedure, the infer-and-synthesise mode of the tool
  provides a way to automatically generate additional useful
  documentation for future users.
%
\item Specifications are integrated into the code as comments for easy
integration with existing tools, as well as
  documentation systems (\eg{}, we support Doxygen and FORD syntax).
%
\item Our specification format is deliberately abstract, with a small
  number of combinators that \emph{do not involve any indexing
    expressions}, \eg{} does not involve expressions like
  \texttt{a(i+1,j-1)}. This contrasts with other stencil specification
  approaches. For example, the work of Kamil et al.~\cite{kamil2016verified}
  provides full-functionality specifications of stencil computations from which
  equivalent DSL code can be generated.  If a user wishes to specify, say, the
  Naiver-Stokes stencil above, they must write a specification which is very
  similar, involving textually dense array-indexing expressions. Therefore any
  low-level mistakes that could be made programming complex indexing code could
  also be made when writing a specification. Our specifications are much more
  abstract, and also more lightweight with the aim of aiding adoption by
  scientists.
\end{itemize}

It should be noted that our verification tool does not deal in the
class of bugs that can be detected automatically (\emph{push-button
  verification}). Instead, stencil computation bugs must largely be
identified relative to a specification of the intended behaviour.

In this paper, we make the following contributions:
%
\begin{itemize}
\item we introduce our specification language for stencils
  that captures many common forms of data access pattern
  (\S~\ref{sec:lang});

\item we detail checking and inference algorithms for our
  specifications (\S~\ref{sec:analysis}), derived from a
  denotational model for the specificaiton language
  (\S~\ref{sec:semantics});

\item we report on a quantitative study of stencil computations on a
  corpus of numerical Fortran programs (ranging from small to large)
  (\S~\ref{sec:evaluation}), totalling one million lines of code. Our tool
  identifies and infers specifications for 35,000 stencil computations in the
  corpus. Approximately 10\% of the stencils we found are non-trivial,
  corresponding to code which is a possible source of errors. This validates
  our hypothesis that the majority of stencil computations have a regular
  shape, and validates the design of our language in its capability to capture
  many core patterns.

\item We give a detailed verification case study of two
  particular stencil computation commonly used in scientifc computing,
  simulating programming errors. Our approach is able to detect all
  possible indexing errors for these computations.
\end{itemize}
%

\bibliography{references}
